Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
reworked the overall paper structure and added annotation on what to do for the next few hours for me

%TODO egb: revisit this paragraph after reading all paper to refine problem statement and solution.
In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it. We mention some of the limitations and how we worked around them. We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.

@@

When the Squeak VM was released it consisted mainly in:
\begin{itemize}
\item an interpreter with a spaghetti stack, 
\item a memory manager with a compact but complex object representation: a pointer-reversing tracing garbage collector and a heap divided into two generations,
\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine,
\item external C code and makefiles to support running the VM on popular platforms.
\end{itemize}

@@

In this paper, we present key details of the Simulation infrastructure, give examples that demonstrate its productivity advantages, and present disadvantages that the architecture presents.
The paper is structured as follows. Section \ref{sec:VMSimulation} introduces the simulation infrastructure used to develop and debug the VM.
%In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM. 
Section \ref{sec:GCExp} and Section \ref{sec:JITExp}  reports our experience developing respectively the full GC and the JIT with our infrastructure. Section \ref{sec:Analysis} shows how we reuse the simulator to analyse the runtime and direct design decisions to improve performance. Lastly, we discuss some related work and conclude.

\section{Virtual Machine Infrastructure}
\label{sec:VMSimulation}

%The main components of \openSmalltalkVM, the interpreter, the JIT and the memory manager, are written in Slang. The platform code, \emph{i.e.,} Operating System dependent code (such as file management or I/O), is written directly in C with a corresponding Smalltalk version for simulation. 
Before detailing the simulation infrastructure used to develop and debug the \openSmalltalkVM, we first briefly introduce the compilation pipeline to generate the production VM.

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code to C code, generating a few files. This first step takes several seconds. Secondly, the C compiler \footnote{depending on the platform, LLVM, GCC, MSVC} translates the C code into an executable. This second step may take a few minutes the first time. But, depending on what the programmer changed and what the C compiler has cached, it usually only takes a few seconds to recompile the C code.

The VM can actually be configured in two main flavours, interpreter-only or interpreter+JIT. Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes \footnote{and on devices that outlaw JITs}. For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding JIT complexity.

@@

\subsection{Simulating the Virtual Machine}

%Simulation allows for manipulating and inspecting Smalltalk image files.
%In general, the simulator provides a toolkit for manipulating and inspecting image files.
%\eem{this feels repetitious.  we should say this once, and once only.  where is it best said?}
The key idea of the Simulation is to allow developers to reuse the whole Smalltalk IDE including the browser, inspectors and debugger to develop the VM. Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.

%In Simulation, one of the core feature is to be able to re-use the whole Smalltalk IDE, including the browser, inspectors and debugger to develop the VM. Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.

As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT. Before detailing the simulation of the interpreter-only VM and the simulation with the JIT, we briefly describe the start-up sequence of \openSmalltalkVM using snapshots, a key aspect of the VM simulator success.

%In this section, we firstly describe briefly the start-up sequence of \openSmalltalkVM using snapshots, a key aspect of the VM simulator success. Then, we detail the simulation of the interpreter-only VM. The simulation of the full VM is a superset and is explained in the last subsection.

\paragraph{Start-up from snapshots.} 
Smalltalk is an object \emph{system}, rather than a language. The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap. 
%A Smalltalk system is started by loading a snapshot file into a VM. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory, swizzling pointers as required, and resumes execution in the process active when the snapshot was created.  
%A new snapshot can be made during or at the end of the session.
When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
More precisely, the snapshot includes objects (such as the classes), the compiled methods in the form of bytecodes and the running processes.
Developing applications consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from objects. Programming may be done live, as the application under development is running. %, for example creating or revising methods in the debugger, and continuing. 
For example, objects may have their shape changed on the fly as instance variables are added and removed.  \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
A new snapshot can be made during or at the end of the development session.

%When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
%Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects. Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.  Objects may have their shape changed on the fly as instance variables are added and removed.  \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.

%\paragraph{Miscellaneous Uses}
%Starting up \openSmalltalkVM from source files has never been supported. However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files. 
%The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.
%The simulator is also used to create 64-bit images from 32-bit images, a transformation that involves replacing certain instances of classes by certain others, for example, the 64-bit system supports an immediate floating point type that represents a subset of 64-bit IEEE 754 floating point numbers whose exponent fits in the same 8-bit range as 32-bit IEEE 754 floats, whereas in the 32-bit system all floats are 8 byte boxed objects.

\paragraph{Interpreter-only simulation.} 
%As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.
 Since executing machine code on the processor simulators is not fast, the simulated interpreter may be faster \footnote{but JIT code is so much more optimized than bytecode interpretation that it may indeed be faster, depending on the code being executed}, and startup time and memory footprint are smaller.

@@

Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part. Let us detail briefly the memory used by the production VM, on the top of the figure. On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM. On higher addresses, we can find the beginning of the memory managed by the VM.

At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects. Usually at very high addresses, we find the C stack. Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone. Both stacks are disjoint and managed differently, the stack zone being broken up into small stack pages \cite{StackMappingVW}.

In the simulator, the heap is stored as a large contiguous byte array. References between objects are indices inside the byte array instead of pointers. All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects. They use specific wrapper classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack zone is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM. In the interpeter configuration the stack zone is a Smalltalk Array object.  In the JIT it is in the byte array, to allow access to the stack from generated machine code.

@@

In addition to the interpreter simulator, simulating the JIT requires simulate the execution of the native code it generates. The JIT itself is written in Slang and simulated wi th the Smalltalk execution model. To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, SMalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely. All but the MIPSLE simulators are accessed through SMalltalk primitives.  These are effectively a combination of an FFI call to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. The next two sections describe in detail how this scheme is used to ensure that simulation and production machine code agree, how simulation machien code is interfaced to the rest of teh simulation, and the debugging facilities for machine code that are available.

@@

\paragraph{JIT fidelity and the interfacing of machine code to the simulation objects}
A key requirement to enable effective development of the JIT is that the code generated during simulation be as close as possible to the code generated by the production JIT.  The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation, a feature we augment with a synthetic clock so that time advances in lock step with code execution. Hence simulations are exactly repeatable.  But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.  The technique we use, involving illegal addresses, also allows us to interface the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and jit, and variables within them, with machine code. To explain this let's consider how the interpreter references a stack frame.

@@

@@

\paragraph{Arbitrary breakpoints, history, and decorated disassembly}
The JIT maintains a simulation-only variable that determines whether the procerssor smulator is invoked in single-stepping mode, or run-until-exception-or-interrupt mode.  When in single stepping mode a circular buffer remembers N previous instructions, allowing one to examine an arbitrary number of instructions (by default 160) preceeding some error.  The arbitrary breakpoint facility is intelligent enough to only enable single stepping once code has been JITted at that address, using run mode until that point.

@@

@@

@@

\subsection{In-image compilation}

@@

To work around these problems, we implemented a tool called \emph{In-image compilation}. In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e. \emph{not} in the simulated heap) to generate the corresponding machine code. In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code and decorate the disassembly and then dumps the output in a text wwindow. Indeed the above snippet is from an in-image compilation.  To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulaiton byte array.  This includes mock addresses for all the object the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.

@@

\paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes. We will use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and  
a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.

@@

\subsection{Virtual Machine Simulation Limitations}

@@

\section{Garbage Collection Development}
\label{sec:GCExp}

Recently, to evaluate new old space garbage collection algorithms we designed against standard algorithms, we implemented a Mark-Sweep in addition to the existing Mark-Compact collector. The whole implementation was done in the simulator, and only when it was working there, was it compiled to C. Using this process, the compiled C code was kind of working out of the box. To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertions. Each assertion ensures a specific state is as the VM developer would expect it to be and stops or log the incorrect result if not. We have a multi-level assertion system. Assertions written in plain Smalltalk are convenient since more complex constraints can be expressed easily using high level structures (Sets, etc.), but they can be performed only in the simulator. Assertions written in Slang are executed in the simulator but they are also compiled to C. The C compiler, based on a compilation flag, chooses to compile the VM with or without assertions. The VM with assertions is used for debugging and to recreate snapshots just before a VM crash. The VM without assertions is the production VM.

\section{Just-in-Time Compiler Development}
\label{sec:JITExp}

We firstly discuss how we use the simulator to debug crashes in deployed applications, then how we use in-image compilation to develop the JIT itself.

\subsection{Debugging crashes with conditional breakpoints in machine code}

\subsection{Optimizing the templates with in-image compilation}

\section{Virtual Machine Analysis: Directing performance decisions}
\label{sec:Analysis}

\subsection{Analysis example}

\subsection{Directing the VM behavior}

Analysing the question is easy in our context.  This analysis took about an hour to perform, including writing an iterator over send sites, and then implementing the eager promotion to open PICs.  Somewhat similar analyses one of the authors performed on the second generation Deutsch Schiffman VM \cite{EfficientST80} , which is written entirely in C, when adding the same PIC scheme took several days, having to be written statically and tested with a traditional edit-compile-debug cycle. The productivity difference is extreme.

\section{Related Work}

@@

@@

\subsection{Conclusion}

>>>>

%TODO egb: revisit this paragraph after reading all paper to refine problem statement and solution.
In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the garbage collector and the just-in-time compiler within it. %We mention some of the limitations and how we worked around them. 
%We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. 
Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.

@@

When the Squeak VM was released it consisted mainly in:
\begin{itemize}
\item an interpreter with a spaghetti stack, 
\item a memory manager with a compact but complex object representation: a pointer-reversing tracing garbage collector and a heap divided into two generations,
\item WarpBlt, a rotation and scaling extension for the bit-based BitBlt graphics engine,
\item external C code and makefiles to support running the VM on popular platforms.
\end{itemize}

@@

In this paper, we present key details of the Simulation infrastructure, give examples that demonstrate its productivity advantages, and present disadvantages that the architecture presents.
The paper is structured as follows. Section \ref{sec:VMSimulation} introduces the simulation infrastructure used to develop and debug the VM.
%In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM. 
Section \label{sec:Exp} reports our experience developing the VM with the simulation infrastructure. Lastly, we discuss some of the simulation limitations, the related work and conclude.

\section{Virtual Machine Simulation}
\label{sec:VMSimulation}

The key idea of the Simulation is to allow developers to reuse the whole Smalltalk IDE including the browser, inspectors and debugger to develop the VM. Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.

%Before detailing the simulation infrastructure used to develop and debug the \openSmalltalkVM, 
We first briefly introduce the compilation pipeline to generate the production VM and Smalltalk snapshots, key elements to understand the design of the simulation infrastructure. Then, we describe how the memory layout of both the production and simulation runtimes. In the following subsection, we explain specific aspects of the simulation infrastructure such as the simulation of the machine code generated by the JIT through the processor simulator. The last subsection discusses features related to the development of the JIT, including for example breakpoints in the machine code it generates.

\subsection{Context}

\paragraph{VM compilation.} 
The main components of \openSmalltalkVM, the interpreter, the JIT and the memory manager, are written in Slang. The platform code, \emph{i.e.,} Operating System dependent code (such as file management or I/O), is written directly in C.

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code to C code, generating a few C files. This first step takes several seconds. Secondly, the C compiler, depending on the platform, LLVM, GCC or MSVC, translates all the C files into an executable. This second step can take up to a few minutes the first time. But, depending on what the programmer changed and what the C compiler has cached, it usually only takes a few seconds to recompile the C code.

The VM can actually be configured in two main flavours, interpreter-only or interpreter+JIT. Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes and on devices that outlaw JITs. For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding JIT complexity.

@@

\paragraph{Snapshots.}
Smalltalk is an object \emph{system}, rather than a language. The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap. 
%A Smalltalk system is started by loading a snapshot file into a VM. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory, swizzling pointers as required, and resumes execution in the process active when the snapshot was created.  
%A new snapshot can be made during or at the end of the session.
When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
More precisely, the snapshot includes objects (such as the classes), the compiled methods in the form of bytecodes and the running processes.
Developing applications consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from objects. Programming may be done live, as the application under development is running. %, for example creating or revising methods in the debugger, and continuing. 
For example, objects may have their shape changed on the fly as instance variables are added and removed%\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
. A new snapshot can be made during or at the end of the development session.

Snapshots can be used to avoid long startup times for specific bugs. The VM can be run to a point where the bug is about to be injected, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the set up time to reach the bug.

\subsection{Memory architectures}

@@

\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part. Let us detail briefly the memory used by the production VM, on the top of the figure. On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.

On higher addresses, we can find the beginning of the memory managed by the VM. At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.

Usually at very high addresses, we find the C stack. Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone. Both stacks are disjoint and managed differently, the stack zone being broken up into small stack pages \cite{StackMappingVW}.

\paragraph{Simulation memory layout.}In the simulator, the heap is stored as a large contiguous byte array. References between objects are indices inside the byte array instead of pointers. All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects. They use specific wrapper classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack zone is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM. In the interpreter-only simulator, the stack zone is a Smalltalk Array object. In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only was implemented first in the easiest way possible, but it did not work out with the JIT.}.

\subsection{Runtime simulation}

\paragraph{Modularity.}

@@

In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of the native code it generates. The JIT itself is written in Slang and simulated wi th the Smalltalk execution model. To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely. All but the MIPSLE simulators are accessed through Smalltalk primitives.  These are effectively a combination of an FFI call to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. The next two sections describe in detail how this scheme is used to ensure that simulation and production machine code agree, how simulation machine code is interfaced to the rest of the simulation, and the debugging facilities for machine code that are available.

@@

\paragraph{Deterministic simulation.}

To be able to reproduce the same bug the exact same multiple times in a row, we designed the simulator to be as deterministic as possible. The most important things to note are the following:
\begin{itemize}
\item Simulated memory is not subject to Address Space Layout Randomisation,
\item A synthetic clock is used so that time advances in lock step with code execution.
\end{itemize}

\paragraph{Interfacing machine code with simulation objects.}

A key requirement to enable effective development of the JIT is that the code generated during simulation be as close as possible to the code generated by the production JIT.  The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation. But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.  The technique we use, involving illegal addresses, also allows us to interface the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and jit, and variables within them, with machine code. To explain this let's consider how the interpreter references a stack frame.

@@

\todo{Shorten and abstract next paragraph. }

@@

\subsection{Machine code debugging}

\paragraph{Instruction recording.}

The JIT maintains a simulation-only variable that determines whether the processor simulator is invoked in single-stepping mode, or run-until-exception-or-interrupt mode.  When in single stepping mode a circular buffer remembers N previous instructions and associated register state, allowing one to examine an arbitrary number of instructions (by default 160) preceding some error. The breakpoint facility is intelligent enough to only enable single stepping once code has been JITted at that address, using run mode until that point.

\paragraph{Conditional breakpoints.}

\todo{Rework. maybe check interaction with section JIT debug.}

@@

\paragraph{In-image compilation.}

\todo{Merge 1 and 2 + some shortening}

%1

@@

\todo{RM end of next paragraph or ref}

@@

%2

@@

To work around these problems, we implemented a tool called \emph{In-image compilation}. In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e. \emph{not} in the simulated heap) to generate the corresponding machine code. In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code and decorate the disassembly and then dumps the output in a text window. Indeed the above snippet is from an in-image compilation.  To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulaiton byte array.  This includes mock addresses for all the object the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.

@@

\paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes. We use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and  
a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.

@@

\section{Experience reports}
\label{sec:Exp}

\todo{add intro about this is anecdote and examples of how we used it ; fix section }

\subsection{Implementing a new full GC algorithm}

\todo{Remove Assertion references}
\todo{Make it more like a story like the following 2 subsec}
Recently, to evaluate new old space garbage collection algorithms we designed against standard algorithms, we implemented a Mark-Sweep in addition to the existing Mark-Compact collector. The whole implementation was done in the simulator, and only when it was working there, was it compiled to C. Using this process, the compiled C code was kind of working out of the box. To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

%\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertions. Each assertion ensures a specific state is as the VM developer would expect it to be and stops or log the incorrect result if not. We have a multi-level assertion system. Assertions written in plain Smalltalk are convenient since more complex constraints can be expressed easily using high level structures (Sets, etc.), but they can be performed only in the simulator. Assertions written in Slang are executed in the simulator but they are also compiled to C. The C compiler, based on a compilation flag, chooses to compile the VM with or without assertions. The VM with assertions is used for debugging and to recreate snapshots just before a VM crash. The VM without assertions is the production VM.

%We firstly discuss how we use the simulator to debug crashes in deployed applications, then how we use in-image compilation to develop the JIT itself.

\subsection{Debugging a crash in machine code with conditional breakpoints}

\subsection{Optimizing the store templates with in-image compilation}

\subsection{Machine code zone send analysis}

\todo{Make following 2 subsections more like a story like the previous ones}

\subsection{Directing VM optimizations through analysis}

Analysing the question is easy in our context.  This analysis took about an hour to perform, including writing an iterator over send sites, and then implementing the eager promotion to open PICs.  Somewhat similar analyses one of the authors performed on the second generation Deutsch Schiffman VM \cite{EfficientST80}, which is written entirely in C, when adding the same PIC scheme took several days, having to be written statically and tested with a traditional edit-compile-debug cycle. The productivity difference is extreme.

\section{Discussion and Related Work}

\subsection{Virtual Machine simulation limitations}

\todo{Merge 1 and 2}

%1
\paragraph{Interpreter-only simulation.} 
 Since executing machine code on the processor simulators is not fast, the simulated interpreter may be faster \footnote{but JIT code is so much more optimized than bytecode interpretation that it may indeed be faster, depending on the code being executed}, and startup time and memory footprint are smaller.

 %2

@@

\subsection{VM Simulator and Bootstrapping facilities}

\subsection{Related Work}

@@

\todo{We need to mention something about graal, maybe high level lang for VM and debugging in general then refs}

@@

\section*{Conclusion}
